---
title: "Final_version"
author: "Jacklyn Peng"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Open file (.tsv)

```{r}
library(readr)

#Gene
ribo_gene_counts <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/salmon.merged.gene_counts.tsv"
polya_gene_counts <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/salmon.merged.gene_counts.tsv"
ribo_gene_lengths <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/salmon.merged.gene_lengths.tsv"
polya_gene_lengths <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/salmon.merged.gene_lengths.tsv"

ribo_gene_counts_data <- read_tsv(ribo_gene_counts, show_col_types = FALSE)
polya_gene_counts_data <- read_tsv(polya_gene_counts, show_col_types = FALSE)
ribo_gene_lengths_data <- read_tsv(ribo_gene_lengths, show_col_types = FALSE)
polya_gene_lengths_data <- read_tsv(polya_gene_lengths, show_col_types = FALSE)

#Transcript
ribo_transcript_counts <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/salmon.merged.transcript_counts.tsv"
polya_transcript_counts <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/salmon.merged.transcript_counts.tsv"
ribo_transcript_lengths <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/salmon.merged.transcript_lengths.tsv"
polya_transcript_lengths <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/salmon.merged.transcript_lengths.tsv"

ribo_transcript_counts_data <- read_tsv(ribo_transcript_counts, show_col_types = FALSE)
polya_transcript_counts_data <- read_tsv(polya_transcript_counts, show_col_types = FALSE)
ribo_transcript_lengths_data <- read_tsv(ribo_transcript_lengths, show_col_types = FALSE)
polya_transcript_lengths_data <- read_tsv(polya_transcript_lengths, show_col_types = FALSE)
  
# Reslut
read_tsv(ribo_gene_counts, n_max = 5)
read_tsv(ribo_gene_lengths, n_max = 5)

read_tsv(ribo_transcript_counts, n_max = 5)
read_tsv(ribo_transcript_lengths, n_max = 5)
```

## Manual calculate Transcripts Per Million (TPM)

```{r}
library(readr)
library(dplyr)

# Function calculate TPM
calculate_tpm <- function(counts, lengths, output_file = NULL) {
  
  # Convert to data frame
  counts <- as.data.frame(counts)
  lengths <- as.data.frame(lengths)
  
  # Convert all columns except gene_id and gene_name to numeric
  counts <- counts %>%
    mutate(across(-c(gene_id, gene_name), as.numeric))
  lengths <- lengths %>%
    mutate(across(-c(gene_id, gene_name), as.numeric))
  
  # # Check for any non-numeric values after conversion (excluding gene_id and gene_name)
  # if (any(sapply(counts[, !names(counts) %in% c("gene_id", "gene_name")], function(x) !is.numeric(x)))) {
  #   stop("Non-numeric values found in counts after conversion. Please check the data.")
  # }
  # if (any(sapply(lengths[, !names(lengths) %in% c("gene_id", "gene_name")], function(x) !is.numeric(x)))) {
  #   stop("Non-numeric values found in lengths after conversion. Please check the data.")
  # }
  
  # Set gene_id as row.names and remove both gene_id and gene_name columns
  rownames(counts) <- counts$gene_id
  counts <- counts[, !names(counts) %in% c("gene_id", "gene_name")]  # Remove both gene_id and gene_name
  rownames(lengths) <- lengths$gene_id
  lengths <- lengths[, !names(lengths) %in% c("gene_id", "gene_name")]  # Remove both gene_id and gene_name

  # Check if dimensions match
  if (!identical(dim(counts), dim(lengths))) {
    stop("Dimensions of counts and lengths do not match!")
  }

  # Check for zero lengths to avoid division by zero
  if (any(lengths == 0, na.rm = TRUE)) {
    warning("Some transcript lengths are zero. These rows will be removed to avoid division by zero.")
    valid_rows <- rowSums(lengths == 0) == 0  # Identify rows with non-zero lengths
    counts <- counts[valid_rows, ]
    lengths <- lengths[valid_rows, ]
  }
  
  # Step 1: Divide counts by transcript lengths to get reads per base
  reads_per_base <- counts / lengths

  # Step 2: Calculate the sum of reads per base for each sample (column)
  col_sums <- colSums(reads_per_base, na.rm = TRUE)

  # Step 3: Normalize by column sums and scale to per million (TPM)
  tpm <- t(t(reads_per_base) / col_sums) * 1e6

  # Optionally save the result to a file
  if (!is.null(output_file)) {
    write.table(tpm, output_file, sep = "\t", quote = FALSE, row.names = TRUE, col.names = NA)
    message("TPM values saved to: ", output_file)
  }

  # Return the TPM data frame
  return(tpm)
}
  
tpm_result <- calculate_tpm(ribo_gene_counts_data, ribo_gene_lengths_data, "salmon_tpm_values.tsv")
head(tpm_result)
```

## Open file (.sf)

```{r}
library(tximport)

files <- c(
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/ribo_URS007_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/ribo_URS008_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/ribo_URS009_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/ribo_URS0010_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/ribo_URS0011_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/ribo_URS0013_quant.sf",
  
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/polya_URS007_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/polya_URS008_quant.sf",  
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/polya_URS009_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/polya_URS0010_quant.sf",  
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/polya_URS0011_quant.sf",
  "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya/polya_URS0013_quant.sf"
)

names(files) <- c(
  "ribo_URS007", "ribo_URS008", "ribo_URS009", "ribo_URS0010", "ribo_URS0011", "ribo_URS0013",
  
  "polya_URS007","polya_URS008", "polya_URS009", "polya_URS0010", "polya_URS0011", "polya_URS0013"
)

file.exists(files)
tx2gene <- read.table("/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo/tx2gene.tsv", 
                      header = TRUE, sep = "\t")
head(tx2gene)

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

head(txi$counts)
head(txi$length)
head(txi$abundance)
```

## Sample Table

```{r}
sampleTable_eABC <- data.frame(
  condition = c("ribo", "ribo", "ribo", "ribo", "ribo", "ribo", 
                "polya", "polya", "polya", "polya", "polya", "polya"),
  row.names = c("ribo_URS007", "ribo_URS008", "ribo_URS009", "ribo_URS0010", "ribo_URS0011", "ribo_URS0013",
                "polya_URS007", "polya_URS008", "polya_URS009", "polya_URS0010", "polya_URS0011", "polya_URS0013"),
  endometriosis = c("No", "Yes", "No", "Yes", "No", "No", 
                    "No", "Yes", "No", "Yes", "No", "No"),
  age_at_biopsy_A = c(29, 38, 35, 36, 34, 35, 
                      29, 38, 35, 36, 34, 35),
  cycle_length_B = c(22, 26, 30, 26.5, 28, 28, 
                     22, 26, 30, 26.5, 28, 28),
  day_of_biopsy_C = c(20, 23, 23, 18, 20, 26, 
                      20, 23, 23, 18, 20, 26),
  patient = c("URS007", "URS008", "URS009", "URS0010", "URS0011", "URS0013", 
              "URS007", "URS008", "URS009", "URS0010", "URS0011", "URS0013"),
  estimated_time = c(68, 77, 69, 77, 78, 70, 
                     68, 77, 69, 77, 78, 70)
)

```

## DESeq2

```{r}
library(DESeq2)

# Filter -> DESeq2
dds <- DESeqDataSetFromTximport(txi, sampleTable_eABC, ~condition)

# Filter low counts genes to reduce noise and improve the power of differential expression analysis
keep <- rowSums(counts(dds)) >= 10 

dds <- dds[keep, ]  # Reset dds 
dds <- DESeq(dds) 

# Result
results <- results(dds, contrast = c("condition", "polya", "ribo"), lfcThreshold = 0.5, alpha = 0.05)

# Summary of results
summary(results)

```

## Global structure and key driver
### Method effect

```{r}
library(ggplot2)
vsd <- vst(dds, blind = TRUE)  # Variance-stabilizing transformation
rlog <- rlog(dds, blind = TRUE) # Regularized Log Transformation

pcaData_condition <- plotPCA(vsd, intgroup = c("condition"), returnData = TRUE)
percentVar_condition <- round(100 * attr(pcaData_condition, "percentVar_condition"))

# Create PCA plot using ggplot2
ggplot(pcaData_condition, aes(x = PC1, y = PC2, color = condition, shape = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_condition[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_condition[2], "% variance")) +
  theme_minimal() +
  coord_fixed()

# Compare ribo_URS0011 with other ribo sample
ribo_counts <- assay(vsd)[, c("ribo_URS007", "ribo_URS008", "ribo_URS009", "ribo_URS0010", "ribo_URS0011", "ribo_URS0013")]
cor(ribo_counts)

# Compare polya_URS0011 with other ribo sample
polya_counts <- assay(vsd)[, c("polya_URS007", "polya_URS008", "polya_URS009", "polya_URS0010", "polya_URS0011", "polya_URS0013")]
cor(polya_counts)
```

### Estimate Uterine Cycle Time

```{r}
library(endest)

# Modified the frame
matrix <- txi$abundance

# Estimate cycle time
results_EUC <- estimate_cycle_time(exprs=matrix)
results_EUC$estimated_time
```

### Estimate Uterine Cycle Time Effect

```{r}
# dds for estimated_time
dds_estimated_time <- DESeqDataSetFromTximport(txi, colData = sampleTable_eABC, 
                                         design = ~ estimated_time)

# Run DESeq2 analysis
dds_estimated_time <- DESeq(dds_estimated_time)

# Variance-stabilizing transformation for PCA
vsd_estimated_time <- vst(dds_estimated_time, blind = TRUE)

# PCA plot with correct intgroup
plotPCA(vsd_estimated_time, intgroup = c("estimated_time"))

# Extract PCA data
pcaData_estimated_time <- plotPCA(vsd_estimated_time, intgroup = c("estimated_time"), returnData = TRUE)

# Calculate the percentage of variance explained by PC1 and PC2
percentVar_estimated_time <- round(100 * attr(pcaData_estimated_time, "percentVar_estimated_time"))

# PCA plot with non-overlapping labels
ggplot(pcaData_estimated_time, aes(x = PC1, y = PC2, color = estimated_time, label = estimated_time)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_estimated_time[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_estimated_time[2], "% variance")) +
  theme_minimal() +
  coord_fixed()

# Quantify estimated_time effect on PC2 using ANOVA
anova_result_estimated_time <- aov(PC2 ~ estimated_time, data = pcaData_estimated_time)
summary(anova_result_estimated_time)

# Calculate correlation between estimated_time and PC2
cor.test(pcaData_estimated_time$PC2, pcaData_estimated_time$estimated_time, method = "pearson") # p-value = 4.82e-11, cor = -0.9942689

```

### Sample Variability

```{r}
library(pheatmap)

# Extract expression data for all samples from vsd
all_counts <- assay(vsd)[, rownames(sampleTable_eABC)]
all_cor <- cor(all_counts)

# Create a data frame for sample annotations (only condition)
annotation_df <- data.frame(
  Condition = sampleTable_eABC$condition,
  row.names = rownames(sampleTable_eABC)
)

# Define colors for annotations (only condition)
annotation_colors <- list(
  Condition = c(ribo = "aquamarine", polya = "deeppink")
)

cor_range <- range(all_cor, na.rm = TRUE)  # Get the actual range of correlation values
min_cor <- max(-1, cor_range[1])  # Ensure the minimum is not less than -1
max_cor <- min(1, cor_range[2])   # Ensure the maximum is not more than 1

# If the range is narrow (e.g., 0.9 to 1.0), focus on that range for finer color differentiation
# For demonstration, let's assume the range is 0.9 to 1.0 (adjust based on actual data)
min_cor <- 0.9  # Adjust this based on your actual data
max_cor <- 1.0  # Adjust this based on your actual data

# Create a finer color gradient with more levels
n_colors <- 100  # Increase the number of color levels for finer granularity
color_gradient <- colorRampPalette(c("blue", "white", "deeppink"))(n_colors)

# Create finer breaks for the color gradient
color_breaks <- seq(min_cor, max_cor, length.out = n_colors + 1)

# Step 4: Draw heatmap with finer color differentiation
pheatmap(
  all_cor,                       # Correlation matrix for all samples
  cluster_rows = TRUE,           # Cluster samples (rows)
  cluster_cols = TRUE,           # Cluster samples (columns)
  annotation_row = annotation_df, # Sample annotations for rows (only condition)
  annotation_col = annotation_df, # Sample annotations for columns (only condition)
  annotation_colors = annotation_colors, # Colors for annotations
  show_rownames = TRUE,          # Show sample names (rows)
  show_colnames = TRUE,          # Show sample names (columns)
  main = "Correlation Heatmap of All Samples (poly-A vs. ribo)", # Title
  fontsize = 10,                 # Font size
  color = color_gradient,        # Finer color gradient
  breaks = color_breaks,         # Finer breaks for the color gradient
  display_numbers = TRUE,        # Display correlation values in the heatmap
  number_color = "black"         # Color of the numbers
)
```

## Genomic region composition (Exon/Intron/Intergenic)

```{r}
# File path
file_list <- list.files(path = "/Users/jacklynpeng/Desktop/Research_Project/R_analysis/region_reads", pattern = "*.csv", full.names = TRUE)

# Process files
result_list <- lapply(file_list, function(file) {
  df <- read.csv(file, skip = 1)
  
  sample_id <- gsub("_reads.csv", "", basename(file))  # Extract ID

  exonic <- sum(df$Tag_count[df$Group %in% c("CDS_Exons", "5' UTR Exons", "3' UTR Exons")])
  intronic <- sum(df$Tag_count[df$Group == "Introns"])
  intergenic <- sum(df$Tag_count[!(df$Group %in% c("CDS_Exons", "5' UTR Exons", "3' UTR Exons", "Introns"))])
  total <- exonic + intronic + intergenic

  data.frame(
    Patient = sample_id,
    Region = c("Exonic", "Intronic", "Intergenic"),
    Percentage = c(exonic, intronic, intergenic) / total * 100
  )
})

# Combine to df
all_results <- bind_rows(result_list)

# ggplot
ggplot(all_results, aes(x = Patient, y = Percentage, fill = Region)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)),
            position = position_stack(vjust = 0.5),
            size = 3, color = "white") +
  labs(title = "Read Distribution by Region",
       y = "Percentage (%)", x = "") +
  scale_fill_manual(values = c(
    "Exonic" = "#F8766D",
    "Intronic" = "#00BFC4",
    "Intergenic" = "#D3D3D3"
  )) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## RNA biotype composition
### RNA biotype (protein_coding, lncRNA, snRNA, ribozyme, and other)

```{r}
library(ggplot2)

# Collect TPM data
tpm_matrix <- txi$abundance
tpm_df <- as.data.frame(tpm_matrix)
tpm_df$gene_id <- rownames(tpm_matrix)

# Make sure biotypes gene ID matchs to tpm_df
colnames(biotypes)[colnames(biotypes) == "ensembl_gene_id"] <- "gene_id"

# Merge tpm and biotypes
tpm_with_biotype <- merge(tpm_df, biotypes[, c("gene_id", "gene_biotype")], by = "gene_id", all.x = TRUE)

# Check gene is not included
if (any(is.na(tpm_with_biotype$gene_biotype))) {
  warning("Some genes do not have a biotype assigned. Number of unmapped genes: ",
          sum(is.na(tpm_with_biotype$gene_biotype)))
}

# If there is not included gene -> unknown
tpm_with_biotype$gene_biotype[is.na(tpm_with_biotype$gene_biotype)] <- "unknown"

# long format
tpm_long <- tpm_with_biotype %>%
  tidyr::pivot_longer(
    cols = matches("^(ribo_|polya_)"),
    names_to = "sample",
    values_to = "tpm"
  )

# Summary biotype by sample
tpm_summary <- tpm_long %>%
  group_by(sample, gene_biotype) %>%
  summarise(tpm = sum(tpm, na.rm = TRUE)) %>%
  ungroup()

# Calculate each sample's tpm
tpm_summary <- tpm_summary %>%
  group_by(sample) %>%
  mutate(total_tpm = sum(tpm)) %>%
  ungroup() %>%
  mutate(ratio = tpm / total_tpm)

tpm_summary$condition <- ifelse(grepl("polya", tpm_summary$sample), "polya", "ribo")

# Choose protein_coding, lncRNA, snRNA, ribozyme, and other
tpm_summary$biotype_group <- dplyr::case_when(
  tpm_summary$gene_biotype == "protein_coding" ~ "protein_coding",
  tpm_summary$gene_biotype == "lncRNA"         ~ "lncRNA",
  TRUE                                         ~ "other"
)

# Definition main biotypes 
main_biotypes <- c("protein_coding", "lncRNA", "snRNA", "ribozyme")
tpm_summary <- tpm_summary %>%
  mutate(
    original_biotype = gene_biotype,  # keep original biotype
    gene_biotype = ifelse(original_biotype %in% main_biotypes,
                          original_biotype, "other")
  )

biotype_colors <- c(
  "protein_coding"   = "#F8766D", 
  "lncRNA"           = "#FCBBA1",  
  "snRNA"            = "#66C2A5",  
  "ribozyme"         = "#00BFC4",  
  "other"            = "#D3D3D3"   
)

p <- ggplot(tpm_summary, aes(x = sample, y = ratio, fill = gene_biotype)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = ifelse(ratio > 0.05, sprintf("%.2f%%", ratio * 100), "")), 
            position = position_stack(vjust = 0.5),
            size = 2, color = "white", fontface = "bold") +
  labs(title = "RNA Biotype Proportions",
       fill = "Biotype",
       y = "Proportion") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = biotype_colors) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, face = "bold"),
    plot.margin = margin(l = 40, r = 40, t = 30, b = 20)
  )

print(p)
```

### Top3 'other' biotypes vs other_minus

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# Step 1. Top 3 original biotypes inside "other" (by total TPM across all samples)
top3_other <- tpm_summary %>%
  filter(!(original_biotype %in% main_biotypes)) %>%
  group_by(original_biotype) %>%
  summarise(total_tpm = sum(tpm, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_tpm)) %>%
  slice_head(n = 3)

# Step 2. Build refined groups (reuse the object you already created)
tpm_summary_other <- tpm_summary %>%
  mutate(
    biotype_refined = case_when(
      original_biotype %in% top3_other$original_biotype ~ original_biotype,
      !(original_biotype %in% main_biotypes) ~ "other_minus",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(biotype_refined))

# Step 3. Denominator = sample total TPM across ALL biotypes
sample_totals <- tpm_summary %>%
  group_by(sample) %>%
  summarise(total_tpm_all = sum(tpm, na.rm = TRUE), .groups = "drop")

# Step 4. Compute proportions for the refined groups
tpm_summary_prop <- tpm_summary_other %>%
  group_by(sample, biotype_refined) %>%
  summarise(total_tpm = sum(tpm, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_totals, by = "sample") %>%
  mutate(prop = total_tpm / total_tpm_all)

# Step 5. Colors (top3 + other_minus)
top3_cols <- brewer.pal(n = 3, "Set2")
names(top3_cols) <- top3_other$original_biotype
biotype_colors <- c("other_minus" = "#D3D3D3", top3_cols)

# Step 6. Plot
ggplot(tpm_summary_prop, aes(x = sample, y = prop, fill = biotype_refined)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = ifelse(prop > 0.005, sprintf("%.2f%%", prop * 100), "")),
            position = position_stack(vjust = 0.5),
            size = 2, color = "white", fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = biotype_colors, name = "Biotype (other group)") +
  labs(x = "Sample", y = "Proportion", title = "Top3 'other' biotypes vs other_minus") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))

```

## Method concordance (per-patient)

```{r}
library(DESeq2)
library(ggplot2)
library(gridExtra)

# Get normalized counts
norm_counts <- counts(dds, normalized = TRUE)

plot_list <- list()
all_outliers <- data.frame()  # (placeholder; not used below)
patients <- unique(sampleTable_eABC$patient)

# Collect all expression values across samples to compute global axis limits
all_polya_expr <- c()
all_ribo_expr  <- c()

for (patient in patients) {
  polya_sample <- which(sampleTable_eABC$patient == patient & sampleTable_eABC$condition == "polya")
  ribo_sample  <- which(sampleTable_eABC$patient == patient & sampleTable_eABC$condition == "ribo")
  
  # 🔑 log2-normalized counts
  polya_expr <- log2(as.numeric(norm_counts[, polya_sample]) + 1)
  ribo_expr  <- log2(as.numeric(norm_counts[, ribo_sample]) + 1)
  
  all_polya_expr <- c(all_polya_expr, polya_expr)
  all_ribo_expr  <- c(all_ribo_expr, ribo_expr)
}

global_x_min <- min(all_ribo_expr, na.rm = TRUE)
global_x_max <- max(all_ribo_expr, na.rm = TRUE)
global_y_min <- min(all_polya_expr, na.rm = TRUE)
global_y_max <- max(all_polya_expr, na.rm = TRUE)

# Plot per patient
for (patient in patients) {
  polya_sample <- which(sampleTable_eABC$patient == patient & sampleTable_eABC$condition == "polya")
  ribo_sample  <- which(sampleTable_eABC$patient == patient & sampleTable_eABC$condition == "ribo")
  
  polya_expr <- log2(as.numeric(norm_counts[, polya_sample]) + 1)
  ribo_expr  <- log2(as.numeric(norm_counts[, ribo_sample]) + 1)
  
  # Pearson correlation
  cor_result <- cor.test(polya_expr, ribo_expr, method = "pearson")
  cor_value <- round(cor_result$estimate, 3)
  p_value   <- format.pval(cor_result$p.value, digits = 3)
  
  df_DE <- data.frame(Polya = polya_expr,
                      Ribo  = ribo_expr,
                      Gene  = rownames(norm_counts))
  
  p <- ggplot(df_DE, aes(x = Ribo, y = Polya)) +
    geom_point(color = "black", alpha = 0.5) +
    geom_smooth(method = "lm", color = "#F8766D", se = FALSE) +
    labs(title = paste("Patient:", patient, "\nCor =", cor_value, ", p =", p_value),
         x = "Ribo Expression (log2 normalized)",
         y = "Polya Expression (log2 normalized)") +
    theme_minimal() +
    xlim(global_x_min, global_x_max) +
    ylim(global_y_min, global_y_max)
  
  plot_list[[as.character(patient)]] <- p
}

# Arrange all patient plots together
combined_plot <- arrangeGrob(grobs = plot_list, nrow = 2, ncol = 3)

# Save
ggsave("Polya_vs_Ribo_log2_normalized.png", combined_plot, width = 15, height = 10, dpi = 300)
ggsave("Polya_vs_Ribo_log2_normalized.pdf", combined_plot, width = 15, height = 10)

```

### All outliers

```{r}
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(gridExtra)

set.seed(1)  # stable label placement

# Get normalized counts (genes x samples)
norm_counts <- counts(dds, normalized = TRUE)

# Patients present in sample table
patients <- sort(unique(sampleTable_eABC$patient))

all_outliers <- list()

# ---------- Per-patient outlier analysis ----------
for (patient in patients) {
  polya_idx <- which(sampleTable_eABC$patient == patient &
                     sampleTable_eABC$condition == "polya")
  ribo_idx  <- which(sampleTable_eABC$patient == patient &
                     sampleTable_eABC$condition == "ribo")
  
  # Skip if either condition missing
  if (length(polya_idx) == 0 || length(ribo_idx) == 0) next
  
  # ✅ Handle replicates correctly: average per gene across columns
  polya_expr <- log2(rowMeans(norm_counts[, polya_idx, drop = FALSE]) + 1)
  ribo_expr  <- log2(rowMeans(norm_counts[, ribo_idx,  drop = FALSE]) + 1)
  
  df_DE <- tibble(
    Patient = patient,
    Gene    = rownames(norm_counts),
    Polya   = polya_expr,
    Ribo    = ribo_expr
  )
  
  # Linear model residuals (standardized)
  fit <- lm(Polya ~ Ribo, data = df_DE)
  res_std <- rstudent(fit)
  df_DE <- df_DE %>%
    mutate(resid_std = res_std,
           resid_abs = abs(resid_std))
  
  # Top 5% residuals = outliers
  threshold <- quantile(df_DE$resid_abs, 0.95, na.rm = TRUE)
  outliers <- df_DE %>%
    filter(resid_abs > threshold) %>%
    mutate(direction = ifelse(resid_std > 0, "Polya>Ribo", "Ribo>Polya"))
  
  all_outliers[[as.character(patient)]] <- outliers
}

all_outliers <- bind_rows(all_outliers)

# ---------- Add gene symbols ----------
all_outliers <- all_outliers %>%
  left_join(
    biotypes %>% dplyr::select(ensembl_gene_id, external_gene_name),
    by = c("Gene" = "ensembl_gene_id")
  ) %>%
  mutate(label = coalesce(external_gene_name, Gene))

# ---------- Plot (all patients' outliers) ----------
# Label only the top 30 per patient to reduce clutter
labels_df <- all_outliers %>%
  group_by(Patient) %>%
  slice_max(order_by = resid_abs, n = 30, with_ties = FALSE) %>%
  ungroup()

p_all <- ggplot(all_outliers, aes(x = Ribo, y = Polya, color = Patient)) +
  geom_point(alpha = 0.4) +
  geom_text_repel(
    data = labels_df,
    aes(label = label),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.3,
    show.legend = FALSE
  ) +
  labs(
    title = "All Outliers Across Patients (Top by standardized residuals)",
    x = "Ribo Expression (log2 normalized)",
    y = "Polya Expression (log2 normalized)"
  ) +
  theme_minimal()

print(p_all)
ggsave("All_outliers_with_gene_names.png", p_all, width = 12, height = 10, dpi = 300)

# ---------- Find genes shared across patients ----------
# Use distinct pairs so a gene counted once per patient
gene_counts <- all_outliers %>%
  distinct(Patient, Gene) %>%
  dplyr::count(Gene, name = "n_patients") %>%
  arrange(desc(n_patients))

shared_genes <- gene_counts %>%
  filter(n_patients >= 2) %>%
  left_join(
    biotypes %>% dplyr::select(ensembl_gene_id, external_gene_name),
    by = c("Gene" = "ensembl_gene_id")
  ) %>%
  relocate(external_gene_name, .after = Gene)

print(head(shared_genes, 20))

# ---------- Export CSV ----------
write.csv(shared_genes, "shared_outlier_genes.csv", row.names = FALSE)

```

## Gene expression concordance (by biotype)
### Gene biotype to 7 classes

```{r}
collapse_biotype_7 <- function(x) {
  x <- as.character(x)
  out <- dplyr::case_when(
    # 1) pseudogene 
    grepl("pseudogene", x, ignore.case = TRUE) ~ "pseudogene",

    # 2) protein_coding（including IG/TR coding）
    x == "protein_coding" |
      grepl("^IG_.*_gene$", x) | grepl("^TR_.*_gene$", x) ~ "protein_coding",

    # 3) lncRNA 
    grepl("lncRNA|lincRNA|antisense|processed_transcript|sense_intronic|sense_overlapping|macro_lncRNA|3prime_overlapping_ncRNA|non[_-]?coding|^TEC$",
          x, ignore.case = TRUE) ~ "lncRNA",

    # 4) snoRNA（including scaRNA）
    grepl("^snoRNA|^scaRNA", x, ignore.case = TRUE) ~ "snoRNA",

    # 5) miRNA
    grepl("miRNA", x, ignore.case = TRUE) ~ "miRNA",

    # 6) rRNA
    grepl("^rRNA$|^Mt_rRNA$|5S_rRNA|5\\.8S_rRNA|12S_rRNA|16S_rRNA|18S_rRNA|28S_rRNA",
          x, ignore.case = TRUE) ~ "rRNA",

    # 7) snRNA
    grepl("^snRNA", x, ignore.case = TRUE) ~ "snRNA",

    TRUE ~ NA_character_
  )
  factor(out, levels = c("lncRNA","protein_coding","pseudogene","snoRNA","miRNA","rRNA","snRNA"))
}
```

### Concordance of gene expression (By biotype)

```{r}
library(edgeR)
library(ggplot2)
library(dplyr)
library(ggrepel)

# counts matrix from DESeq2 object
counts_matrix <- counts(dds)

# Compute CPM (non-log)
cpm_matrix <- cpm(counts_matrix, log = FALSE)

# ---- Select polya / ribo sample columns from  sample table ----
ribo_samples  <- rownames(sampleTable_eABC)[sampleTable_eABC$condition == "ribo"]
polya_samples <- rownames(sampleTable_eABC)[sampleTable_eABC$condition == "polya"]

# Ensure names line up with the matrix columns
ribo_samples  <- intersect(colnames(cpm_matrix), ribo_samples)
polya_samples <- intersect(colnames(cpm_matrix), polya_samples)

if (length(ribo_samples) == 0 || length(polya_samples) == 0) {
  stop("No matched ribo/polya columns found. Check that colnames(dds) match rownames(sampleTable_eABC).")
}

# ---- Average expression across samples for each condition ----
ribo_mean  <- log2(rowMeans(cpm_matrix[, ribo_samples,  drop = FALSE]) + 1)
polya_mean <- log2(rowMeans(cpm_matrix[, polya_samples, drop = FALSE]) + 1)

df <- data.frame(
  gene_id = rownames(counts_matrix),
  ribo    = ribo_mean,
  polya   = polya_mean,
  stringsAsFactors = FALSE
)

# ---- Join annotations and collapse biotypes into 7 classes ----
annotation_df <- biotypes %>%
  dplyr::rename(
    gene_id          = ensembl_gene_id,
    gene_name        = external_gene_name,
    gene_biotype_raw = gene_biotype
  ) %>%
  distinct(gene_id, .keep_all = TRUE)

# Remove Ensembl version suffix like ".12"
df$gene_id <- sub("\\.\\d+$", "", df$gene_id)
annotation_df$gene_id <- sub("\\.\\d+$", "", annotation_df$gene_id)

df <- df %>% left_join(annotation_df, by = "gene_id")

df$gene_type <- collapse_biotype_7(df$gene_biotype_raw)

# ---- Flag outliers and compute per-biotype Pearson correlations ----
df$highlight <- abs(df$polya - df$ribo) > 4

# Correlation per biotype + position of the r label in each facet (top-left with a small inset)
cors <- df %>%
  filter(!is.na(gene_type)) %>%
  group_by(gene_type) %>%
  summarise(r = cor(ribo, polya, method = "pearson", use = "complete.obs"),
            .groups = "drop") %>%
  left_join(
    df %>%
      filter(!is.na(gene_type)) %>%
      group_by(gene_type) %>%
      summarise(
        xmin = min(ribo,  na.rm = TRUE),
        xmax = max(ribo,  na.rm = TRUE),
        ymin = min(polya, na.rm = TRUE),
        ymax = max(polya, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "gene_type"
  ) %>%
  mutate(
    label = paste0("r = ", round(r, 3)),
    x = xmin + 0.02 * (xmax - xmin),
    y = ymax - 0.02 * (ymax - ymin)
  )

# Top-5 outliers per biotype by absolute polya - ribo difference
outliers <- df %>%
  filter(!is.na(gene_type)) %>%
  group_by(gene_type) %>%
  arrange(desc(abs(polya - ribo))) %>%
  slice_head(n = 5) %>%
  ungroup()

# ---- Plot: facet by gene_type (rows), compare ribo vs polya ----
p <- ggplot(df %>% filter(!is.na(gene_type)),
            aes(x = ribo, y = polya)) +
  geom_point(aes(color = highlight), alpha = 1, size = 1.2) +
  scale_color_manual(values = c("#D3D3D3", "#F8766D"), labels = c("No", "Yes")) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid") +
  facet_grid(gene_type ~ ., scales = "free") +
  theme_bw(base_size = 13) +
  labs(x = "Ribo log2(CPM+1)", y = "PolyA log2(CPM+1)", color = "|Δ| > 4") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text  = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  ) +
  ggrepel::geom_text_repel(
    data = outliers,
    aes(label = gene_name),
    size = 3,
    color = "#F8766D",
    max.overlaps = 15,
    min.segment.length = 0
  ) +
  geom_text(
    data = cors,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 0, vjust = 1, size = 4
  )

print(p)
ggsave("enodometrium.png", p, width = 10, height = 14, dpi = 300)

```

### All outliers (Appendix A)

```{r}
library(dplyr)

# Prepare deltas and labels
df <- df %>%
  mutate(
    delta      = polya - ribo,
    abs_delta  = abs(delta),
    gene_label = ifelse(!is.na(gene_name) & gene_name != "", gene_name, gene_id)
  )

# Select all outliers (highlight == TRUE), ordered within each gene_type by |delta| (no rank column)
outliers_by_subtype_endometrium <- df %>%
  filter(!is.na(gene_type), highlight) %>%  # if highlight may be NA, use coalesce(highlight, FALSE)
  group_by(gene_type) %>%
  arrange(desc(abs_delta), .by_group = TRUE) %>%
  ungroup() %>%
  transmute(
    gene_type,
    gene = gene_label,
    gene_id,
    ribo  = round(ribo, 3),
    polya = round(polya, 3),
    delta = round(delta, 3),
    abs_delta = round(abs_delta, 3)
  )

# Print all rows (set n to a smaller number if the table is huge)
print(outliers_by_subtype_endometrium, n = Inf)

# Write CSV output
write.csv(outliers_by_subtype_endometrium, "outliers_by_subtype_endometrium.csv", row.names = FALSE)

```

## Differential expression: gene- and biotype-level
### DEA (polya_riched and ribo_riched)

```{r}
# Remove NA
valid_results <- results[!is.na(results$log2FoldChange) & !is.na(results$padj), ]

# Filter out polya_riched_genes
polya_riched_genes <- rownames(results[valid_results$log2FoldChange > 0.5 & valid_results$padj < 0.05, ])
head(polya_riched_genes)

# Filter out ribo_riched_genes
ribo_riched_genes <- rownames(results[valid_results$log2FoldChange < -0.5 & valid_results$padj < 0.05, ])
head(ribo_riched_genes)

# Check NA
sum(!is.na(results))  # 184857
sum(!is.na(polya_riched_genes))  # 3088
sum(!is.na(ribo_riched_genes)) # 1140
sum(!is.na(valid_results)) # 169014
```

### Global DE (MA plot)

```{r}
#sig_genes_MA <- subset(valid_results, padj < 0.1)
sig_genes_MA <- subset(valid_results, padj < 0.05)
head(sig_genes_MA)

library(ggplot2)

ma_df <- as.data.frame(valid_results)

ggplot(ma_df, aes(x = log10(baseMean + 1), y = log2FoldChange,
                  color = padj < 0.05)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "#F8766D")) +
  theme_minimal() +
  labs(x = "log10(baseMean+1)", y = "log2FC",
       title = "MA plot")

```

### Volcano & significance

```{r}
library(ggplot2)
library(ggrepel)

# Assuming valid_results is your DESeq2 results
Volcano_df <- as.data.frame(valid_results)
Volcano_df$ensembl_gene_id <- rownames(Volcano_df)
Volcano_df <- merge(Volcano_df, 
                    biotypes[, c("ensembl_gene_id", "external_gene_name")],
                    by = "ensembl_gene_id", 
                    all.x = TRUE)

# Filter more significant genes with padj < 0.01 and abs(log2FoldChange) > 2
sig_genes_Volcano <- subset(Volcano_df, padj < 0.05 & abs(log2FoldChange) > 2)
head(sig_genes_Volcano)

# Volcano plot with text labels for significant genes
ggplot(Volcano_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = padj < 0.05)) +
  labs(title = "Volcano Plot") +
  scale_color_manual(values = c("black", "#F8766D")) +
  geom_text_repel(
    data = sig_genes_Volcano,
    aes(label = external_gene_name),
    box.padding = 0.5, 
    max.overlaps = 15
  )
```

###  Biotype-stratified volcano plots (protein-coding, lncRNA, pseudogene)

```{r}
library(edgeR)
library(ggplot2)
library(dplyr)
library(forcats)

# ===== thresholds/labels =====
alpha   <- 0.05
lfc_thr <- 1
gene_levels <- c("protein_coding","lncRNA","pseudogene", "snoRNA", "rRNA", "snRNA")

strip_ver <- function(x) sub("\\.\\d+$", "", x)

# Keep only the intersection of samples and put them in the order of sampleTable_eABC
common_samples <- intersect(colnames(counts_matrix), rownames(sampleTable_eABC))
counts_matrix <- counts_matrix[, common_samples, drop = FALSE]
sampleTable_eABC <- sampleTable_eABC[common_samples, , drop = FALSE]

# ===== Grouping (baseline = ribo) =====
grp <- factor(sampleTable_eABC$condition, levels = c("ribo", "polya"))

# ===== EdgeR QLF (logFC = polya − ribo) =====
keep <- edgeR::filterByExpr(counts_matrix, group = grp)

y <- edgeR::DGEList(counts = counts_matrix[keep, ], group = grp)
y <- edgeR::calcNormFactors(y, method = "TMM")

design <- model.matrix(~ grp)                  # (Intercept) + grppolya
y <- edgeR::estimateDisp(y, design, robust = TRUE)
fit <- edgeR::glmQLFit(y, design, robust = TRUE)

qlf <- edgeR::glmQLFTest(fit, coef = 2)        # polya vs ribo
de  <- edgeR::topTags(qlf, n = Inf)$table
de$gene_id <- rownames(de)

# ===== Annotate =====
de$gene_id <- strip_ver(de$gene_id)

annotation_df <- biotypes %>%
  dplyr::rename(
    gene_id          = ensembl_gene_id,
    gene_name        = external_gene_name,
    gene_biotype_raw = gene_biotype
  ) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE)

annotation_df$gene_id <- strip_ver(annotation_df$gene_id)

de_annot <- de %>%
  dplyr::left_join(annotation_df, by = "gene_id") %>%
  dplyr::mutate(
    gene_type    = collapse_biotype_7(gene_biotype_raw),
    gene_type    = forcats::fct_drop(factor(as.character(gene_type), levels = gene_levels)),
    neglog10FDR  = -log10(pmax(FDR, .Machine$double.eps)),
    sig          = (FDR < alpha) & (abs(logFC) >= lfc_thr),
    Direction    = dplyr::case_when(
      sig & logFC > 0 ~ "polya-enriched",
      sig & logFC < 0 ~ "ribo-enriched",
      TRUE            ~ "NS"
    ),
    Direction    = factor(Direction, levels = c("ribo-enriched","NS","polya-enriched"))
  ) %>%
  dplyr::filter(!is.na(gene_type))

# ===== Volcano plot (facet by gene_type only) =====
pA <- ggplot(de_annot, aes(x = logFC, y = neglog10FDR)) +
  geom_point(aes(color = Direction), size = 0.7, alpha = 0.85, na.rm = TRUE) +
  geom_vline(xintercept = c(-lfc_thr, lfc_thr), linetype = "dashed") +
  geom_hline(yintercept = -log10(alpha), linetype = "dashed") +
  facet_grid(. ~ gene_type) +
  scale_color_manual(
    values = c("ribo-enriched" = "#00BFC4", "NS" = "grey70", "polya-enriched" = "#F8766D"),
    breaks = c("ribo-enriched","polya-enriched"),
    name = "Direction"
  ) +
  labs(
    x = "log2FC (polya vs ribo)",
    y = "-log10(adj.Pvalue)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position   = "right",
    strip.background  = element_rect(fill = "grey90", colour = NA),
    strip.text        = element_text(face = "bold")
  )

print(pA)
ggsave("Volcano_DE_by_endometrium.png", pA, width = 14, height = 5.5, dpi = 300)

```

#### Fraction of biotype-stratified volcano plots (protein-coding, lncRNA, pseudogene)

```{r}
library(dplyr)
library(ggplot2)
library(forcats)

sum_C <- de_annot %>%
  filter(sig) %>%
  mutate(
    Direction_label = dplyr::case_when(
      Direction == "ribo-enriched" ~ "ribo-enriched",
      Direction == "polya-enriched"   ~ "polya-enriched",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Direction_label)) %>%
  group_by(gene_type, Direction_label) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(de_total = sum(n), pct = 100 * n / de_total) %>%
  ungroup() %>%
  mutate(
    Direction_label = factor(Direction_label,
                             levels = c("ribo-enriched", "polya-enriched")),
    gene_type = fct_drop(gene_type)
  )

pC <- ggplot(sum_C, aes(x = gene_type, y = pct, fill = Direction_label)) +
  geom_col(width = 0.3) +
  scale_fill_manual(values = c("ribo-enriched" = "#00BFC4",
                               "polya-enriched"  = "#F8766D"),
                    name = "Direction") +
  labs(x = NULL, y = "Percentage of DE genes (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text.x = element_text(hjust = 0.5)
  )

print(pC)
ggsave("Fraction_DE_endometrium.png", pC, width = 8, height = 6.5, dpi = 300) 

```

### edgeR Ql (snRNA, snoRNA, pseudigene, and rRNA)

```{r}
## ===== Libraries =====
suppressPackageStartupMessages({
  library(edgeR)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(forcats)
})

## ===== Parameters & helpers =====
alpha       <- 0.05
gene_levels <- c("lncRNA","protein_coding","pseudogene","snoRNA","miRNA","rRNA","snRNA")

# Remove Ensembl gene version suffix like ".1"
strip_ver <- function(x) sub("\\.\\d+$", "", x)

# Collapse many Ensembl biotypes into 7 broad classes
collapse_biotype_7 <- function(x) {
  x <- tolower(x)
  out <- ifelse(x %in% c("protein_coding"), "protein_coding",
         ifelse(grepl("pseudogene", x), "pseudogene",
         ifelse(grepl("sno", x), "snoRNA",
         ifelse(grepl("snrna", x), "snRNA",
         ifelse(grepl("mirna", x), "miRNA",
         ifelse(grepl("rrna", x), "rRNA",
         ifelse(grepl("lncrna|linc|antisense|sense_intronic|sense_overlapping|processed_transcript|3prime_overlapping_ncrna|non_coding", x),
                "lncRNA", NA_character_)))))))
  out
}

# Keep the intersection and enforce the same sample order in both objects
counts_matrix    <- counts_matrix[, common_samples, drop = FALSE]
sampleTable_eABC <- sampleTable_eABC[common_samples, , drop = FALSE]

# Define groups (baseline = ribo)
grp <- factor(sampleTable_eABC$condition, levels = c("ribo","polya"))

## ===== Annotation to biotypes =====
ann0 <- biotypes %>%
  dplyr::rename(gene_id = ensembl_gene_id,
         gene_name = external_gene_name,
         gene_biotype_raw = gene_biotype) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  mutate(gene_id  = strip_ver(gene_id),
         gene_type = collapse_biotype_7(gene_biotype_raw))

# Align annotation rows to counts
rownames(counts_matrix) <- strip_ver(rownames(counts_matrix))
ann <- ann0[match(rownames(counts_matrix), ann0$gene_id), c("gene_id","gene_type")]
keep_rows <- !is.na(ann$gene_type)

## ===== A) Biotype-level differential proportion (offset = total library size) =====
# Aggregate counts by biotype (sum across genes within each biotype)
agg <- rowsum(counts_matrix[keep_rows, , drop = FALSE], group = ann$gene_type[keep_rows])

# Keep only biotypes present and in the desired order
present <- intersect(gene_levels, rownames(agg))
agg <- agg[present, , drop = FALSE]

# Use total library size of each sample as the denominator (offset)
total_lib <- colSums(counts_matrix)

dge <- DGEList(counts = agg, group = grp)
dge$samples$lib.size     <- total_lib
dge$samples$norm.factors <- 1  # comparing proportions; leave as 1 for clarity

design <- model.matrix(~ grp)  # coefficient 2 is grppolya (polya vs ribo)
dge <- estimateDisp(dge, design, robust = TRUE)
fit <- glmQLFit(dge, design, robust = TRUE)
qlf <- glmQLFTest(fit, coef = 2)
tab <- topTags(qlf, n = Inf)$table
tab$biotype <- rownames(tab)

# Derive SE from the QL F-statistic: for a single coefficient, F = (beta/SE)^2
tab$F[is.na(tab$F)] <- .Machine$double.eps
se <- abs(tab$logFC) / sqrt(pmax(tab$F, .Machine$double.eps))
z  <- 1.96  # Normal approximation for 95% CI (use qt for t-based CI if you prefer)
tab$lwr <- tab$logFC - z * se
tab$upr <- tab$logFC + z * se

plot_df <- tab %>%
  filter(biotype %in% present) %>%
  mutate(biotype = factor(biotype, levels = rev(present))) %>%
  arrange(biotype)

p_biotype <- ggplot(plot_df, aes(x = biotype, y = logFC)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2) +
  geom_point(size = 3) +
  coord_flip() +
  labs(x = NULL,
       y = "log2FC of biotype proportion (polya vs ribo)",
       title = "Biotype-level differential proportion (edgeR QL)") +
  theme_bw(base_size = 12)

ggsave("FigA_biotype_forest.png", p_biotype, width = 6, height = 4.5, dpi = 300)
print(p_biotype)

## ===== Focus only on snRNA / snoRNA / pseudogene / rRNA =====
focus <- c("snRNA","snoRNA","pseudogene","rRNA")
p_biotype_focus <- p_biotype + scale_x_discrete(limits = rev(intersect(focus, present)), drop = FALSE)
ggsave("FigA_biotype_forest_focus.png", p_biotype_focus, width = 5, height = 3.8, dpi = 300)
print(p_biotype_focus)
```

### Fraction of DE genes by biotype

```{r}
library(ggplot2)
library(dplyr)
library(RColorBrewer)

polya_riched_biotypes <- tpm_with_biotype[tpm_with_biotype$gene_id %in% polya_riched_genes, ]
ribo_riched_biotypes  <- tpm_with_biotype[tpm_with_biotype$gene_id %in% ribo_riched_genes, ]

# Proportion by biotype (tag with condition)
polya_riched_biotype_prop <- polya_riched_biotypes %>%
  filter(!is.na(gene_biotype)) %>%
  dplyr::count(gene_biotype) %>%
  mutate(ratio = n / sum(n), condition = "polya")

ribo_riched_biotype_prop <- ribo_riched_biotypes %>%
  filter(!is.na(gene_biotype)) %>%
  dplyr::count(gene_biotype) %>%
  mutate(ratio = n / sum(n), condition = "ribo")

# Combine
combined_prop <- bind_rows(polya_riched_biotype_prop, ribo_riched_biotype_prop)

# Consistent stacking order across both bars (by overall abundance)
levels_global <- combined_prop %>%
  group_by(gene_biotype) %>%
  summarise(total_ratio = sum(ratio), .groups = "drop") %>%
  arrange(desc(total_ratio)) %>%
  pull(gene_biotype)

combined_prop <- combined_prop %>%
  mutate(
    condition    = factor(condition, levels = c("polya","ribo")),
    gene_biotype = factor(gene_biotype, levels = levels_global)
  )

# Colors
biotype_colors_v2 <- c(
  "protein_coding"       = "#F8766D",
  "IG_gene"              = "#E41A1C",
  "lncRNA"               = "#FCBBA1",
  "antisense"            = "#FDBF6F",
  "processed_transcript" = "#FF7F00",
  "snRNA"                = "#66C2A5",
  "snoRNA"               = "#1F78B4",
  "miRNA"                = "#00BFC4",
  "ribozyme"             = "#00CED1",
  "pseudogene"           = "#999999",
  "other_minus"          = "#D3D3D3"
)

# Auto-extend palette for unseen biotypes
missing_biotypes <- setdiff(levels_global, names(biotype_colors_v2))
if (length(missing_biotypes) > 0) {
  extra_cols <- colorRampPalette(brewer.pal(8, "Set3"))(length(missing_biotypes))
  names(extra_cols) <- missing_biotypes
  biotype_colors_v2 <- c(biotype_colors_v2, extra_cols)
}

# Top-3 labels per condition
top3_labels <- combined_prop %>%
  group_by(condition) %>%
  slice_max(order_by = ratio, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(label = paste0(gene_biotype, " (", round(ratio * 100, 1), "%)"))

# Plot
ggplot(combined_prop, aes(x = condition, y = ratio * 100, fill = gene_biotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = biotype_colors_v2, name = "Gene Biotype") +
  labs(title = "Biotype Proportion: Polya vs Ribo", y = "Proportion (%)", x = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title  = element_text(hjust = 0.2, face = "bold"),
    legend.text = element_text(size = 8),
    legend.title= element_text(face = "bold")
  ) +
  geom_text(
    data = top3_labels,
    aes(label = label),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 1
  )

```

## Gene overlap & discovery
### Overlap (Venn)

```{r}
library(ggVennDiagram)
library(ggplot2)

counts <- DESeq2::counts(dds, normalized = TRUE)

# Collect condition
polya_samples <- which(sampleTable_eABC$condition == "polya")
ribo_samples <- which(sampleTable_eABC$condition == "ribo")

# Collect polya and ribo data
polya_counts <- counts[, polya_samples]
ribo_counts <- counts[, ribo_samples]

# Gene detected more than 0
polya_detected <- rowSums(polya_counts > 0) > 0
ribo_detected <- rowSums(ribo_counts > 0) > 0

# Count gene number
polya_only <- sum(polya_detected & !ribo_detected)
ribo_only <- sum(!polya_detected & ribo_detected)
both_detected <- sum(polya_detected & ribo_detected)

venn_data <- list(
  polya = which(polya_detected),
  ribo = which(ribo_detected)
)

# Create Venn diagram with customized boundaries
ggVennDiagram(venn_data, 
              category.names = c("Poly-A", "Ribo"),
              label_size = 4,
              set_size = 4) +
  scale_fill_gradient(low = "#F8766D", high = "skyblue") +
  ggtitle("Venn Diagram of Genes Detected by Poly-A and Ribo") +
  theme(
    text = element_text(size = 10),
    plot.margin = margin(t = , r = 20, b = 20, l = 20, unit = "pt")  # Adjust margins
  )
```

### Gene Ontology (GO)

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

polya_riched <- enrichGO(gene = upregulated_genes, 
                  OrgDb = org.Hs.eg.db, 
                  keyType = "ENSEMBL",
                  ont = "ALL", 
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05)

ribo_riched <- enrichGO(gene = downregulated_genes, 
                    OrgDb = org.Hs.eg.db, 
                    keyType = "ENSEMBL",
                    ont = "ALL", 
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.1)

head(polya_riched)
head(ribo_riched)

require(DOSE)
dotplot(ribo_riched, showCategory = 10, split = "ONTOLOGY") + 
  facet_grid(~ONTOLOGY)
dotplot(ribo_riched, showCategory=10)


```

## RNA splicing analysis

```{r}
library(Rsubread)
library(edgeR)
library(dplyr)

## ===================== User settings =====================
base_dir <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow"   # <-- Set to your project root; contains polya_bam/ and ribo_bam/ subfolders
gtf      <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/Human_Reference/Homo_sapiens.GRCh38.114.gtf" # <-- Set to your GTF

isPE           <- TRUE   # paired-end?
strandSpecific <- 0      # 0=unstranded, 1=fr-firststrand, 2=fr-secondstrand
n_threads      <- 8

## ===================== Build BAM paths =====================
# Choose subfolder based on condition
folder_for <- function(cond) if (cond == "polya") "polya_bam" else "ribo_bam"

# Given (cond, patient), build candidate filenames and return the first that exists
candidate_paths <- function(cond, patient) {
  dir <- file.path(base_dir, folder_for(cond))
  c(
    file.path(dir, paste0(patient, ".markdup.sorted.bam")),
    file.path(dir, paste0(patient, ".sorted.bam")),
    file.path(dir, paste0(patient, ".bam"))
  )
}
find_bam <- function(cond, patient) {
  hits <- candidate_paths(cond, patient)
  ex   <- file.exists(hits)
  if (any(ex)) hits[which(ex)[1]] else NA_character_
}

sampleTable_eABC$bam <- mapply(find_bam,
                               cond = sampleTable_eABC$condition,
                               patient = sampleTable_eABC$patient,
                               USE.NAMES = FALSE)

# Check for missing BAMs
missing <- which(is.na(sampleTable_eABC$bam))
if (length(missing) > 0) {
  cat("Could not find BAM for samples:\n",
      paste0("  - ", rownames(sampleTable_eABC)[missing], " (", 
             sampleTable_eABC$condition[missing], ", ", sampleTable_eABC$patient[missing], ")\n"),
      sep = "")
  stop("Please adjust 'base_dir' or filename patterns.")
}

cat("Example mapping:\n")
print(head(sampleTable_eABC[, c("condition","patient","bam")], 6))

## ===================== featureCounts (exon level) =====================
grp <- factor(sampleTable_eABC$condition, levels = c("ribo","polya"))

fc <- featureCounts(
  files                 = sampleTable_eABC$bam,
  annot.ext             = gtf,
  isGTFAnnotationFile   = TRUE,
  GTF.featureType       = "exon",
  GTF.attrType          = "gene_id",
  GTF.attrType.extra    = c("exon_id","exon_number"),
  useMetaFeatures       = FALSE,      # per-exon counts (not per-gene)
  isPairedEnd           = isPE,
  strandSpecific        = strandSpecific,
  countMultiMappingReads= FALSE,
  minMQS                = 10,
  nthreads              = n_threads,
#  jcounts               = TRUE        # also count junctions for QC
)

exon.counts <- fc$counts
ann         <- as.data.frame(fc$annotation)

# Stable exon_id; if GTF lacks exon_id, fall back to genomic coordinates
exon_id <- if ("exon_id" %in% names(ann) && !all(is.na(ann$exon_id))) ann$exon_id else
  paste0(ann$GeneID, ":", ann$Chr, ":", ann$Start, "-", ann$End, "(", ann$Strand, ")")

rownames(exon.counts) <- exon_id
gene_id <- ann$GeneID

## ===================== edgeR DEU (~ patient + grp) =====================
y <- DGEList(exon.counts,
             genes = data.frame(gene_id = gene_id, exon_id = rownames(exon.counts)))
keep <- filterByExpr(y, group = grp)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y, method = "TMM")

design <- model.matrix(~ factor(sampleTable_eABC$patient) + grp)  # paired
colnames(design)
coef_idx <- which(colnames(design) == "grppolya")                  # effect of polyA vs ribo

y <- estimateDisp(y, design)
fit <- glmFit(y, design)

s <- diffSpliceDGE(fit, geneid = y$genes$gene_id, exonid = y$genes$exon_id, coef = coef_idx)

# Gene-level: Simes combination of exon p-values
res_genes <- topSpliceDGE(s, test = "Simes", n = Inf)

# Gene-level (alternative F-style test)
res_genes_F <- topSpliceDGE(s, test = "gene", n = Inf)

# Exon-level results (per exon)
res_exons <- topSpliceDGE(s, test = "exon", n = Inf)

# Save and quick peek
write.csv(res_genes,  "DEU_top_genes_Simes.csv", row.names = FALSE)
write.csv(res_genes_F,"DEU_top_genes_geneF.csv", row.names = FALSE)
write.csv(res_exons,  "DEU_top_exons.csv",      row.names = FALSE)

head(res_genes[order(res_genes$FDR), ])
head(res_exons[order(res_exons$FDR), ])


cat("Saved:\n  DEU_top_genes_paired.csv\n  DEU_top_exons_paired.csv\n")

# Visualize one significant gene
if ("plotSpliceDGE" %in% getNamespaceExports("edgeR")) {
  sig_gene <- res_genes$GeneID[which.min(res_genes$FDR)] |> as.character()
  if (length(sig_gene) && !is.na(sig_gene)) {
    png("DEU_plot_example.png", width = 1400, height = 900, res = 150)
    edgeR::plotSpliceDGE(s, geneid = sig_gene)   # <-- no 'coefficient' here
    dev.off()
    cat("Saved example plot: DEU_plot_example.png (gene:", sig_gene, ")\n")
  }
}

```

### Run rMATS using BAMs

```{r}
# Note: ribo-depletion and poly(A) selection differ in intronic coverage, 3' bias, and transcript abundance.
# These can influence exon-coverage–based metrics. To avoid mistaking coverage differences for splicing
# differences, prefer junction-centric detection and thresholds (e.g., ≥20 junction reads per group).
suppressPackageStartupMessages({
  library(fs); library(stringr); library(readr)
  library(processx); library(data.table); library(dplyr); library(ggplot2)
})

# ====== EDIT THESE ======
# Paths (your data are under nextflow)
ribo_bam_dir <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/ribo_bam"
polya_bam_dir <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/polya_bam"
gtf_path <- "/Users/jacklynpeng/Desktop/Research_Project/nextflow/Human_Reference/Homo_sapiens.GRCh38.114.gtf"

# rMATS environment (conda env bin; use ~/micromamba/envs/rmats/bin if you used micromamba)
rmats_env_bin <- "/opt/anaconda3/envs/rmats/bin"
python_exe <- file.path(rmats_env_bin, "python")
rmats_py   <- file.path(rmats_env_bin, "rmats.py")

# Library strandedness
library_type <- "fr-firststrand"  # fr-unstranded / fr-firststrand / fr-secondstrand

# Output / temp folders
rmats_out <- "rmats_out"
rmats_tmp <- "rmats_tmp"
dir_create(rmats_out, recurse = TRUE)
dir_create(rmats_tmp, recurse = TRUE)

# ====== 1) Scan BAMs and write b1/b2 ======
list_bams <- function(dir){
  stopifnot(dir_exists(dir))
  b <- dir_ls(dir, glob = "**/*.bam", recurse = TRUE, type = "file")
  if (!length(b)) stop("No .bam files found in ", dir)
  sorted <- grepl("sorted", basename(b), ignore.case = TRUE)
  b <- c(b[sorted], b[!sorted])
  normalizePath(unique(b), winslash = "/", mustWork = TRUE)
}
bams_ribo  <- list_bams(ribo_bam_dir)
bams_polya <- list_bams(polya_bam_dir)

b1_file <- "b1_ribo.txt"
b2_file <- "b2_polya.txt"
writeLines(paste(bams_ribo,  collapse = ","), b1_file)
writeLines(paste(bams_polya, collapse = ","), b2_file)

cat("\n# Wrote list files:\n",
    b1_file, ": ", length(bams_ribo),  " BAMs\n",
    b2_file, ": ", length(bams_polya), " BAMs\n", sep = "")

# ====== 2) Estimate read length (use 101 if Rsamtools is unavailable) ======
read_length <- 101L
if (requireNamespace("Rsamtools", quietly = TRUE)) {
  bam_try <- bams_ribo[1]
  param <- Rsamtools::ScanBamParam(what = c("qwidth"))
  qw <- tryCatch(Rsamtools::scanBam(bam_try, param = param, yieldSize = 5000)[[1]]$qwidth,
                 error = function(e) integer())
  qw <- qw[is.finite(qw)]
  if (length(qw)) read_length <- as.integer(median(qw))
}
cat(sprintf("# Estimated read length = %d\n", read_length))

# ====== 3) Check executables/files ======
stopifnot(file.exists(python_exe), file.exists(rmats_py), file.exists(gtf_path))

# ====== 4) Clean old tmp & run rMATS (add --variable-read-length) ======
dir_delete(rmats_tmp); dir_create(rmats_tmp)

cmd <- c(
  rmats_py,
  "--b1", b1_file,
  "--b2", b2_file,
  "--gtf", gtf_path,
  "--od", rmats_out,
  "--tmp", rmats_tmp,
  "--readLength", as.character(read_length), "--variable-read-length",
  "--libType", library_type,
  "--nthread", "8"
)

env_px <- c(PATH = paste(rmats_env_bin, Sys.getenv("PATH"), sep = .Platform$path.sep))
cat("\n# Running rMATS (processx)...\n")
processx::run(python_exe, cmd, env = env_px, echo = TRUE, error_on_status = TRUE)
cat("# rMATS completed.\n")

# ====== 5) List outputs, load JCEC, plot (robust) ======
library(fs); library(data.table); library(dplyr); library(ggplot2)

# 1) Inspect output folder tree
message("# rmats_out tree:")
try({ print(fs::dir_tree(rmats_out, recurse = 1)) }, silent = TRUE)

# 2) Use lenient matching to find all MATS files (avoid regex misses)
outs <- fs::dir_ls(rmats_out, glob = "*MATS*.txt", recurse = TRUE)
print(outs)

if (!length(outs)) {
  # Still none found: print summary to diagnose where it stopped
  sumf <- file.path(rmats_out, "summary.txt")
  if (file.exists(sumf)) {
    cat("\n# summary.txt (前 50 行):\n")
    cat(readLines(sumf, n = 50), sep = "\n")
  }
  stop("找不到任何 *MATS*.txt。可能只跑到 count 階段或 b1/b2 沒有複本。")
}

# 3) Prefer JCEC; fall back to JC if JCEC not present
has_jcec <- grepl("\\.MATS\\.JCEC\\.txt$", outs)
if (any(has_jcec)) {
  wanted <- outs[has_jcec]
} else {
  wanted <- outs[grepl("\\.MATS\\.JC\\.txt$", outs)]
}
print(wanted)

# 4) Read the five event types (skip missing types)
read_ev_path <- function(ev, paths) {
  f <- paths[grepl(paste0("^", rmats_out, "/?", ev, "\\.MATS\\.(JCEC|JC)\\.txt$"), paths)]
  if (length(f) == 0) return(NULL)
  data.table::fread(f[1])[, EventType := ev]
}

rmats_raw <- data.table::rbindlist(
  lapply(c("SE","RI","A5SS","A3SS","MXE"), read_ev_path, paths = wanted),
  use.names = TRUE, fill = TRUE
)

# 5) Deduplicate column names (avoid duplicated ID); rename the first ID to EventID
data.table::setnames(rmats_raw, make.unique(names(rmats_raw), sep = "..."))
if ("ID" %in% names(rmats_raw)) data.table::setnames(rmats_raw, "ID", "EventID")

tbl <- rmats_raw |>
  as_tibble() |>
  mutate(
    DeltaPSI = suppressWarnings(as.numeric(IncLevelDifference)),
    FDR      = suppressWarnings(as.numeric(FDR)),
    sig      = is.finite(FDR) & FDR <= 0.05 &
               is.finite(DeltaPSI) & abs(DeltaPSI) >= 0.1
  )

message("# rows: ", nrow(tbl), " | significant: ", sum(tbl$sig, na.rm = TRUE))

# 6) Plot
dir_create("figures")
p_vol <- ggplot(tbl, aes(DeltaPSI, -log10(FDR))) +
  geom_point(aes(shape = EventType, alpha = sig)) +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(x = "ΔPSI (ribo − polya)", y = "-log10(FDR)", title = "rMATS: Differential splicing")
ggsave("figures/fig1a_volcano_rmats.pdf", p_vol, width = 6.5, height = 4.8)

p_cnt <- tbl |> filter(sig) |> dplyr::count(EventType) |>
  ggplot(aes(EventType, n)) + geom_col() +
  labs(y = "# significant events", x = "Event type",
       title = "Significant events (FDR ≤ 0.05 & |ΔPSI| ≥ 0.1)")
ggsave("figures/fig1b_event_type_counts.pdf", p_cnt, width = 5.2, height = 3.8)

```

### |ΔPSI| distribution

```{r}
library(dplyr); library(ggplot2)
df_eff <- tbl %>% filter(sig) %>%
  mutate(direction = ifelse(DeltaPSI > 0, "ribo↑", "polya↑"),
         abs_dpsi = abs(DeltaPSI))

p_violin <- ggplot(df_eff, aes(EventType, abs_dpsi)) +
  geom_violin(trim = FALSE, alpha = 0.3) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  facet_wrap(~direction) +
  labs(x = "Event type", y = "|ΔPSI|",
       title = "Effect size distribution of significant events by type & direction") +
  theme_bw()
ggsave("figures/fig_type_effectsize_violin.pdf", p_violin, width = 7.2, height = 4.2)

```

### Direction enrichment per event type (Fisher's exact test)

```{r}
library(dplyr); library(broom); library(ggplot2); library(purrr); library(tidyr)

df_sig <- tbl %>% filter(sig) %>%
  mutate(direction = ifelse(DeltaPSI > 0, "ribo_up", "polya_up"))

# Build each EventType's 2x2 table
types <- sort(unique(df_sig$EventType))
all_res <- map_dfr(types, function(tp){
  a <- sum(df_sig$EventType == tp & df_sig$direction == "ribo_up")
  b <- sum(df_sig$EventType == tp & df_sig$direction == "polya_up")
  c <- sum(df_sig$EventType != tp & df_sig$direction == "ribo_up")
  d <- sum(df_sig$EventType != tp & df_sig$direction == "polya_up")
  mat <- matrix(c(a,b,c,d), nrow=2, byrow=TRUE,
                dimnames=list(type=c(tp,"others"), dir=c("ribo_up","polya_up")))
  ft <- fisher.test(mat)   # two-sided

  # Haldane-Anscombe correction
  a2<-a+0.5; b2<-b+0.5; c2<-c+0.5; d2<-d+0.5
  or  <- (a2/b2)/(c2/d2)
  se  <- sqrt(1/a2 + 1/b2 + 1/c2 + 1/d2)
  lcl <- exp(log(or) - 1.96*se); ucl <- exp(log(or) + 1.96*se)

  tibble(EventType = tp, a, b, c, d,
         OR = or, CI_low = lcl, CI_high = ucl,
         p_value = ft$p.value)
})

all_res <- all_res %>% mutate(p_adj = p.adjust(p_value, method = "BH"),
                              label = ifelse(OR>1,"ribo↑","polya↑"))

p_forest <- ggplot(all_res,
  aes(y = reorder(EventType, OR), x = log2(OR), xmin = log2(CI_low), xmax = log2(CI_high),
      color = label)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbarh(height = 0) +
  geom_point(size = 2) +
  scale_x_continuous(name = "log2(OR)  ( >0 ⇒ ribo enriched )") +
  labs(y = "Event type", title = "Direction enrichment per event type (Fisher’s exact test)") +
  theme_bw()
ggsave("figures/fig_type_direction_enrichment_forest.pdf", p_forest, width = 6.2, height = 4.5)

# Table
readr::write_tsv(all_res, "figures/type_direction_fisher.tsv")

```

### Direction preference per type (proportion of significant events)

```{r}
library(dplyr); library(ggplot2)

plot_df <- tbl %>%
  dplyr::filter(sig) %>%
  dplyr::mutate(direction = ifelse(DeltaPSI > 0, "ribo↑", "polya↑"))

# (A) Stack
p_stack <- ggplot(plot_df, aes(x = EventType, fill = direction)) +
  geom_bar(position = "stack") +
  labs(x = "Event type", y = "# significant events",
       title = "Significant events by type & direction (ribo↑ = ΔPSI>0)") +
  theme_bw()
ggsave("figures/fig_type_by_direction_counts.pdf", p_stack, width = 6, height = 4)

# (B) Percentage
p_prop <- ggplot(plot_df, aes(x = EventType, fill = direction)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Event type", y = "Proportion within type",
       title = "Direction preference per type (proportion of significant events)") +
  theme_bw()
ggsave("figures/fig_type_by_direction_proportion.pdf", p_prop, width = 6, height = 4)

```

## Appendix A. Additional covariate analyses
### A.1 Patient effect

```{r}
library(ggplot2)

# Calculate the percentage of variance explained by PC1 and PC2
percentVar_patient <- round(100 * attr(pcaData_patient, "percentVar_patient"))

# Create PCA plot using ggplot2
ggplot(pcaData_patient, aes(x = PC1, y = PC2, color = condition, shape = patient)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_patient[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_patient[2], "% variance")) +
  theme_minimal() +
  coord_fixed()
```

###  A.2 Endometriosis status (Yes/No)

```{r}
# dds for endometriosis
dds_endometriosis <- DESeqDataSetFromTximport(txi, 
                                        colData = sampleTable_eABC, 
                                        design = ~ endometriosis)

# Run DESeq2 analysis
dds_endometriosis <- DESeq(dds_endometriosis)

# Variance-stabilizing transformation for PCA
vsd_endometriosis <- vst(dds_endometriosis, blind = TRUE)

# PCA plot with correct intgroup
plotPCA(vsd_endometriosis, intgroup = c("endometriosis"))

# Extract PCA data
pcaData_endometriosis <- plotPCA(vsd_endometriosis, intgroup = c("endometriosis"), returnData = TRUE)

# Calculate the percentage of variance explained by PC1 and PC2
percentVar_endometriosis <- round(100 * attr(pcaData_endometriosis, "percentVar_endometriosis"))

# Create PCA plot using ggplot2
ggplot(pcaData_endometriosis, aes(x = PC1, y = PC2, color = endometriosis, shape = endometriosis)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_endometriosis[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_endometriosis[2], "% variance")) +
  theme_minimal() +
  coord_fixed()

# Quantify endometriosis effect on PC2 using ANOVA
anova_result_endometriosis <- aov(PC2 ~ endometriosis, data = pcaData_endometriosis)
summary(anova_result_endometriosis)

```

###  A.3 Day of biopsy / cycle length ratio

```{r}
# Calculate cycle_length_B / day_of_biopsy_C, build BC
sampleTable_eABC$CB <- sampleTable_eABC$day_of_biopsy_C / sampleTable_eABC$cycle_length_B

# dds for CB
dds_CB <- DESeqDataSetFromTximport(txi, 
                                        colData = sampleTable_eABC, 
                                        design = ~ CB)

# Run DESeq2 analysis
dds_CB <- DESeq(dds_CB)

# Variance-stabilizing transformation for PCA
vsd_CB <- vst(dds_CB, blind = TRUE)

# PCA plot with correct intgroup
plotPCA(vsd_CB, intgroup = c("CB"))

# Extract PCA data
pcaData_CB <- plotPCA(vsd_CB, intgroup = c("CB"), returnData = TRUE)

# Calculate the percentage of variance explained by PC1 and PC2
percentVar_CB <- round(100 * attr(pcaData_CB, "percentVar_CB"))

# PCA plot with non-overlapping labels
ggplot(pcaData_CB, aes(x = PC1, y = PC2, color = CB, label = name)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = Inf) +  # key line: avoids overlap
  xlab(paste0("PC1: ", percentVar_CB[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_CB[2], "% variance")) +
  theme_minimal() +
  coord_fixed()

# Extract PCA data
pcaData_CB <- plotPCA(vsd_CB, intgroup = c("CB"), returnData = TRUE)

# Quantify CB effect on PC2 using ANOVA
anova_result_CB <- aov(PC2 ~ CB, data = pcaData_CB)
summary(anova_result_CB)

# Calculate correlation between CB and PC2
cor.test(pcaData_CB$PC2, pcaData_CB$CB, method = "pearson") # p-value = 0.04979, cor = 0.5764095 
```


## Appendix B. Replication on blood and colon using the same analysis settings
###  B.2 Main findings consistent with prior literature

### In endometrium.Rmd

```{r}
library(dplyr)

# Adjustable label for the dataset; output files will use this name
dataset_label <- "Endometrium"

dir.create("out", showWarnings = FALSE, recursive = TRUE)

# Safety checks
stopifnot(all(c("ribo","polya","gene_type") %in% names(df)))

# If a 'material' column exists, include it in the grouping automatically
group_vars <- c("gene_type")
if ("material" %in% names(df)) group_vars <- c("material", group_vars)

cors <- df %>%
  filter(!is.na(gene_type)) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(
    r_pearson  = cor(ribo, polya, method = "pearson",  use = "complete.obs"),
    r_spearman = cor(ribo, polya, method = "spearman", use = "complete.obs"),
    n          = sum(stats::complete.cases(ribo, polya)),
    .groups = "drop"
  ) %>%
  # Optional: add p-value for Pearson correlation (closed-form via t-statistic)
  mutate(
    t_stat  = ifelse(n >= 3 & abs(r_pearson) < 1, r_pearson * sqrt((n - 2) / (1 - r_pearson^2)), NA_real_),
    p_value = ifelse(n >= 3 & is.finite(t_stat), 2 * stats::pt(-abs(t_stat), df = n - 2), NA_real_)
  ) %>%
  mutate(dataset = dataset_label) %>%
  relocate(dataset)

# Save both RDS and CSV
stub <- file.path("out", paste0(gsub("\\s+", "_", tolower(dataset_label)), "_cors"))
saveRDS(cors, paste0(stub, ".rds"))
write.csv(cors, paste0(stub, ".csv"), row.names = FALSE)

cors

```

### In blood_colon.Rmd

```{r}
library(dplyr)

dir.create("out", showWarnings = FALSE, recursive = TRUE)

# Safety checks
stopifnot(all(c("ribo","polya","gene_type","material") %in% names(df)))

cors <- df %>%
  filter(!is.na(gene_type), !is.na(material)) %>%
  group_by(material, gene_type) %>%
  summarise(
    r_pearson  = stats::cor(ribo, polya, method = "pearson",  use = "complete.obs"),
    r_spearman = stats::cor(ribo, polya, method = "spearman", use = "complete.obs"),
    n          = sum(stats::complete.cases(ribo, polya)),
    .groups = "drop"
  ) %>%
  # p-value for Pearson via t-statistic (requires n >= 3 and |r| < 1)
  mutate(
    t_stat  = ifelse(n >= 3 & abs(r_pearson) < 1, r_pearson * sqrt((n - 2) / (1 - r_pearson^2)), NA_real_),
    p_value = ifelse(is.finite(t_stat), 2 * stats::pt(-abs(t_stat), df = pmax(n - 2, 1)), NA_real_)
  ) %>%
  mutate(
    dataset = case_when(
      tolower(material) == "blood" ~ "Blood",
      tolower(material) == "colon" ~ "Colon",
      TRUE ~ material
    )
  ) %>%
  dplyr::select(dataset, gene_type, r_pearson, r_spearman, n, p_value)

# Save both RDS and CSV with a consistent stub
stub <- file.path("out", "blood_colon_cors")
saveRDS(cors, paste0(stub, ".rds"))
write.csv(cors, paste0(stub, ".csv"), row.names = FALSE)

cors


```

## Compare Endometrium vs Blood+Colon outliers per biotype

```{r}
## Run Endometrium vs Colon and Endometrium vs Blood in a loop
library(dplyr)
library(stringr)

# --- helpers ---
`%||%` <- function(a, b) if (!is.null(a)) a else b
strip_ver <- function(x) sub("\\.\\d+$", "", x %||% "")

prep_tbl <- function(df, dataset_label) {
  df %>%
    mutate(
      dataset        = dataset_label,
      gene_id_clean  = ifelse(is.na(gene_id) | gene_id == "", NA, strip_ver(gene_id)),
      gene_name_norm = ifelse(is.na(gene)   | gene   == "", NA, toupper(trimws(gene))),
      direction      = dplyr::case_when(
        delta >  0 ~ "PolyA > Ribo",
        delta <  0 ~ "PolyA < Ribo",
        TRUE        ~ "No change"
      )
    )
}

endo_path <- "outliers_by_subtype_endometrium.csv"
bc_path   <- "outliers_by_suptype_blood_and_colon.csv"

if (!file.exists(endo_path)) stop("Missing: ", endo_path)
if (!file.exists(bc_path))   stop("Missing: ", bc_path)

dir.create("out", showWarnings = FALSE, recursive = TRUE)

datasets <- c("Colon", "Blood")

for (target_dataset in datasets) {
  message("\n=== Processing: Endometrium vs ", target_dataset, " ===")

  endo_raw <- read.csv(endo_path, stringsAsFactors = FALSE, check.names = FALSE)
  bc_raw   <- read.csv(bc_path,   stringsAsFactors = FALSE, check.names = FALSE)

  # If the Blood/Colon CSV includes a 'material' column, filter to the chosen tissue
  if ("material" %in% names(bc_raw)) {
    bc_raw <- bc_raw %>%
      mutate(material = tolower(trimws(material))) %>%
      filter(material == tolower(target_dataset))
    if (nrow(bc_raw) == 0) {
      warning("No rows found for material = ", target_dataset,
              " in '", bc_path, "'. Skipping this dataset.")
      next
    }
  } else {
    message("No 'material' column in ", bc_path,
            ". Assuming the file is already specific to: ", target_dataset)
  }

  # Required columns (rank removed)
  needed_cols <- c("gene_type","gene","gene_id","ribo","polya","delta","abs_delta")
  miss_endo <- setdiff(needed_cols, names(endo_raw))
  miss_bc   <- setdiff(needed_cols, names(bc_raw))
  if (length(miss_endo)) {
    warning("Endometrium CSV missing columns: ", paste(miss_endo, collapse = ", "), ". Skipping.")
    next
  }
  if (length(miss_bc)) {
    warning("Second CSV (", target_dataset, ") missing columns: ", paste(miss_bc, collapse = ", "), ". Skipping.")
    next
  }

  endo <- prep_tbl(endo_raw, "Endometrium")
  bc   <- prep_tbl(bc_raw,   target_dataset)

  # --- Primary match by Ensembl gene_id (version stripped) ---
  id_matches <- endo %>%
    filter(!is.na(gene_id_clean)) %>%
    inner_join(
      bc %>% filter(!is.na(gene_id_clean)),
      by = "gene_id_clean",
      suffix = c("_endo", "_bc")
    ) %>%
    mutate(match_type = "gene_id")

  # --- Secondary match by gene name for those NOT matched by ID ---
  endo_unmatched <- endo %>%
    anti_join(dplyr::select(id_matches, gene_id_clean) %>% distinct(), by = "gene_id_clean")

  bc_unmatched <- bc %>%
    anti_join(dplyr::select(id_matches, gene_id_clean) %>% distinct(), by = "gene_id_clean")

  name_matches <- endo_unmatched %>%
    filter(!is.na(gene_name_norm)) %>%
    inner_join(
      bc_unmatched %>% filter(!is.na(gene_name_norm)),
      by = "gene_name_norm",
      suffix = c("_endo", "_bc")
    ) %>%
    mutate(match_type = "gene_name")

  # --- Combine matches, compute comparison fields (no rank) ---
  overlap <- dplyr::bind_rows(id_matches, name_matches) %>%
    transmute(
      match_type,
      gene_id         = dplyr::coalesce(gene_id_clean_endo, gene_id_clean_bc),
      gene_name       = dplyr::coalesce(gene_endo,          gene_bc),
      gene_type_endo  = gene_type_endo,
      gene_type_bc    = gene_type_bc,
      delta_endo      = delta_endo,
      delta_bc        = delta_bc,
      abs_delta_endo  = abs_delta_endo,
      abs_delta_bc    = abs_delta_bc,
      direction_endo  = direction_endo,
      direction_bc    = direction_bc,
      same_biotype    = gene_type_endo == gene_type_bc,
      same_direction  = sign(delta_endo) == sign(delta_bc)
    )

  # --- Summary ---
  summary_counts <- list(
    dataset_2               = target_dataset,
    n_endo                  = nrow(endo),
    n_second                = nrow(bc),
    n_matches_by_id         = nrow(id_matches),
    n_matches_by_name       = nrow(name_matches),
    n_matches_total         = nrow(overlap),
    n_same_direction        = sum(overlap$same_direction, na.rm = TRUE),
    pct_same_direction      = round(100 * mean(overlap$same_direction, na.rm = TRUE), 1),
    n_same_biotype          = sum(overlap$same_biotype,   na.rm = TRUE),
    pct_same_biotype        = round(100 * mean(overlap$same_biotype,   na.rm = TRUE), 1)
  )

  cat("\n=== OVERLAP SUMMARY (Endometrium vs ", target_dataset, ") ===\n", sep = "")
  print(summary_counts)

  cat("\n=== Overlap by biotype (counts) ===\n")
  overlap_by_biotype <- overlap %>%
    group_by(gene_type_endo, gene_type_bc) %>%
    summarise(n = n(), .groups = "drop") %>%
    arrange(desc(n))
  print(overlap_by_biotype, n = 100)

  cat("\n=== Direction consistency by biotype ===\n")
  dir_consistency <- overlap %>%
    mutate(biotype_pair = paste0(gene_type_endo, " | ", gene_type_bc)) %>%
    group_by(biotype_pair) %>%
    summarise(
      matches = n(),
      same_dir = sum(same_direction, na.rm = TRUE),
      pct_same_dir = round(100 * mean(same_direction, na.rm = TRUE), 1),
      .groups = "drop"
    ) %>%
    arrange(desc(matches))
  print(dir_consistency, n = 100)

  # --- Unique to each dataset (not matched by ID or name) ---
  endo_unique <- endo %>%
    anti_join(dplyr::select(overlap, gene_id) %>% distinct(), by = c("gene_id_clean" = "gene_id")) %>%
    anti_join(dplyr::select(name_matches, gene_name_norm) %>% distinct(),
              by = c("gene_name_norm" = "gene_name_norm")) %>%
    transmute(
      dataset,
      gene_id = gene_id_clean,
      gene    = gene,
      gene_type,
      delta,
      abs_delta,
      direction
    )

  bc_unique <- bc %>%
    anti_join(dplyr::select(overlap, gene_id) %>% distinct(), by = c("gene_id_clean" = "gene_id")) %>%
    anti_join(dplyr::select(name_matches, gene_name_norm) %>% distinct(),
              by = c("gene_name_norm" = "gene_name_norm")) %>%
    transmute(
      dataset,
      gene_id = gene_id_clean,
      gene    = gene,
      gene_type,
      delta,
      abs_delta,
      direction
    )

  # --- Save outputs ---
  slug <- gsub("\\s+", "_", tolower(target_dataset))
  write.csv(overlap,     paste0("out/endometrium_vs_", slug, "_overlap.csv"), row.names = FALSE)
  write.csv(endo_unique, paste0("out/endometrium_unique_outliers_vs_", slug, ".csv"), row.names = FALSE)
  write.csv(bc_unique,   paste0("out/", slug, "_unique_outliers.csv"), row.names = FALSE)

  cat("\nFiles saved in ./out/:\n",
      paste0("- endometrium_vs_", slug, "_overlap.csv\n"),
      paste0("- endometrium_unique_outliers_vs_", slug, ".csv\n"),
      paste0("- ", slug, "_unique_outliers.csv\n"))
}

```

### Open file

```{r}
library(dplyr)
library(ggplot2)
library(forcats)

ds <- "Colon"
slug <- gsub("\\s+", "_", tolower(ds))

overlap_path <- paste0("out/endometrium_vs_", slug, "_overlap.csv")
endo_u_path  <- paste0("out/endometrium_unique_outliers_vs_", slug, ".csv")
ds_u_path    <- paste0("out/", slug, "_unique_outliers.csv")

overlap    <- read.csv(overlap_path, stringsAsFactors = FALSE, check.names = FALSE)
endo_unique<- read.csv(endo_u_path,  stringsAsFactors = FALSE, check.names = FALSE)
ds_unique  <- read.csv(ds_u_path,    stringsAsFactors = FALSE, check.names = FALSE)

```

### Agreement scatter

```{r}
library(dplyr)
library(ggplot2)

# Ensure a stable mapping: treat logical as an ordered factor so colors don't flip
overlap <- overlap %>%
  dplyr::mutate(same_direction = factor(same_direction, levels = c(TRUE, FALSE)))

# Fixed palette: TRUE is always blue, FALSE is always orange
my_cols <- c(`TRUE` = "#00BFC4", `FALSE` = "#F8766D")

p_scatter <- overlap %>%
  ggplot(aes(x = delta_endo, y = delta_bc, color = same_direction)) +
  geom_hline(yintercept = 0, linetype = "dashed") +      # horizontal zero guide
  geom_vline(xintercept = 0, linetype = "dashed") +      # vertical zero guide
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +  # y = x reference
  geom_point(alpha = 0.6) +                              # points with slight transparency
  coord_equal() +                                        # 1:1 aspect ratio
  labs(
    title = paste("Delta agreement: Endometrium vs", ds),
    x = "delta (Endometrium)",
    y = paste0("delta (", ds, ")"),
    color = "Same direction?"
  ) +
  # Keep TRUE the same color even when FALSE is absent; color NAs light grey
  scale_color_manual(values = my_cols, drop = FALSE, na.value = "grey80") +
  theme_bw(base_size = 12)

print(p_scatter)

# Save to disk at 300 dpi (suitable for slides/print)
ggsave(paste0("out/scatter_delta_endo_vs_", slug, ".png"),
       p_scatter, width = 7, height = 6, dpi = 300)

```

